---
title: 'Assignment 1: Examining multivariate data'
author: "Mohsen Pirmoradiyan, Ahmed Alhasan, Asad Enver, Ali Etminan, Mubarak Hussain"
date: "11/20/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Describing individual variables

Consider the data set in the T1-9.dat le, National track records for women. For 55 different
countries we have the national records for 7 variables (100; 200; 400; 800; 1500; 3000m and
marathon ). Use R to do the following analyses.

a) Describe the 7 variables with mean values, standard deviations e.t.c.

```{r include=FALSE}
library(ggplot2)
library(psych)
```


```{r echo=FALSE, fig.align='center', fig.height=3}


X <- read.table("D:/Workshop/R/Multivariate Statistics/Data/T1-9.DAT")
colnames(X) <- c("Country", "100m", "200m", "400m", "800m", "1500m", "3000m", "Marathon")

Minimum               <- matrix(0,7,1)
colnames(Minimum)     <- c("Min.")

Median                <- matrix(0,7,1)
colnames(Median)      <- c("Median")

Mean                  <- matrix(0,7,1)
colnames(Mean)        <- c("Mean")

S_Deviation           <- matrix(0,7,1)
colnames(S_Deviation) <- c("SD")

Maximum               <- matrix(0,7,1)
colnames(Maximum)     <- c("Max.")


M <- function(X,var) {
  for(i in 1:var) {
    Minimum[i,]     <- min(X[,i+1])
    Median[i,]      <- median(X[,i+1])
    Mean[i,]        <- mean(X[,i+1])
    S_Deviation[i,] <- sd(X[,i+1])
    Maximum[i,]     <- max(X[,i+1])
  }
  result            <- round(cbind(Minimum, Median, Mean, S_Deviation, Maximum),2)
  rownames(result)  <- c("100m", "200m", "400m", "800m", "1500m", "3000m", "Marathon")
  return(result)
}
knitr::kable(M(X = X, var = 7))

```


b) Illustrate the variables with different graphs (explore what plotting possibilities R has). Make
sure that the graphs look attractive (it is absolutely necessary to look at the labels, font
sizes, point types). Are there any apparent extreme values? Do the variables seem normally
distributed? Plot the best fitting (match the mean and standard deviation, i.e. method of
moments) Gaussian density curve on the data's histogram. For the last part you may be
interested in the hist() and density() functions.

```{r echo=FALSE, fig.align='center', fig.height=4.5, fig.width=8}
pairs.panels(X)
```

```{r echo=FALSE, fig.align='center', fig.width=6, fig.height=2.5}

ggplot(X, aes(x=X$`100m`)) +
  geom_histogram(aes(y=..density..), binwidth=.15,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 100m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`200m`)) +
  geom_histogram(aes(y=..density..), binwidth=.3,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 200m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`400m`)) +
  geom_histogram(aes(y=..density..), binwidth=.9,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 400m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`800m`)) +
  geom_histogram(aes(y=..density..), binwidth=.03,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 800m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`1500m`)) +
  geom_histogram(aes(y=..density..), binwidth=.12,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 1500m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`3000m`)) +
  geom_histogram(aes(y=..density..), binwidth=.4,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 3000m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=Marathon)) +
  geom_histogram(aes(y=..density..), binwidth=6,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's Marathon Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

```

### Analysis:

- Looking at the scatterplots we can easily identify 3 to 5 outliers (countries) for each variable. As the race gets longer it is more easier to identify these outliers.

- The first two variables are good approximation to normal distribution, however it starts to be skewed from the third one and the longer the race the more right skewd it will get (more like gamma distribution)


## Question 2: Relationships between the variables

a) Compute the covariance and correlation matrices for the 7 variables. Is there any apparent
structure in them? Save these matrices for future use.

```{r echo=FALSE}


Corr = round(cor(X[,-1]),2)
cat("Correlation Matrix:\n")

knitr::kable(Corr)

cat("Covariance Matrix:\n")
Cov = round(cov(X[,-1]),2)
knitr::kable(Cov)
```

## Analysis:

- There is a structure within the correlation data, it can be identified that there is high correlation between shorter races and also high correlation between longer races but the shorter and longer races are less correlated.

&nbsp;

b) Generate and study the scatterplots between each pair of variables. Any extreme values?

```{r include=FALSE}
library(ggrepel)
library("gridExtra")
```


```{r echo=FALSE, fig.align='center', fig.height=10, fig.width=12}
p1 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,3]))+
  xlab("100m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p2 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,4]))+
  xlab("100m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p3 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,5]))+
  xlab("100m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p4 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,6]))+
  xlab("100m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p5 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,7]))+
  xlab("100m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p6 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,8]))+
  xlab("100m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

grid.arrange(p1, p2, p3, p4, p5, p6)
```


```{r echo=FALSE, fig.align='center', fig.height=12, fig.width=10}


p7 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,2]))+
  xlab("200m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p8 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,4]))+
  xlab("200m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p9 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,5]))+
  xlab("200m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p10 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,6]))+
  xlab("200m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p11 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,7]))+
  xlab("200m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p12 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,8]))+
  xlab("200m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p13 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,2]))+
  xlab("400m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p14 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,3]))+
  xlab("400m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p15 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,5]))+
  xlab("400m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p16 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,6]))+
  xlab("400m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p17 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,7]))+
  xlab("400m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p18 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,8]))+
  xlab("400m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p19 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,2]))+
  xlab("800m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p20 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,3]))+
  xlab("800m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p21 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,4]))+
  xlab("800m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p22 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,6]))+
  xlab("800m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p23 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,7]))+
  xlab("800m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p24 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,8]))+
  xlab("800m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p25 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,2]))+
  xlab("1500m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p26 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,3]))+
  xlab("1500m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p27 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,4]))+
  xlab("1500m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p28 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,5]))+
  xlab("1500m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p29 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,7]))+
  xlab("1500m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p30 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,8]))+
  xlab("1500m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p31 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,2]))+
  xlab("30000m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p32 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,3]))+
  xlab("3000m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p33 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,4]))+
  xlab("3000m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p34 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,5]))+
  xlab("3000m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p35 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,6]))+
  xlab("3000m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p36 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,8]))+
  xlab("3000m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p37 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,2]))+
  xlab("Marathon") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p38 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,3]))+
  xlab("Marathon") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p39 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,4]))+
  xlab("Marathon") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p40 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,5]))+
  xlab("Marathon") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p41 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,6]))+
  xlab("Marathon") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p42 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,7]))+
  xlab("Marathon") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')


grid.arrange(p7, p8, p9, p10, p11, p12)
grid.arrange(p13, p14, p15, p16, p17, p18) 
grid.arrange(p19, p20, p21, p22, p23, p24) 
grid.arrange(p25, p26, p27, p28, p29, p30) 
grid.arrange(p31, p32, p33, p34, p35, p36) 
grid.arrange(p37, p38, p39, p40, p41, p42) 
```


## Analysis:

- Yes, especially when we compare marathon & 3000m races with the shorter races.



&nbsp;

c) Explore what other plotting possibilities R offers for multivariate data. Present other (at
least two) graphs that you find interesting with respect to this data set.


```{r include=FALSE}
library(car)
```


```{r echo=FALSE}
par(mar= c(1,1,1,1))

scatterplotMatrix(X[,2:8])
```

```{r include=FALSE}
library(corrplot)
```


```{r echo=FALSE}

corrplot(Corr, type = "upper", 
        tl.col = "black", tl.srt = 45)

```


## Question 3: Examining for extreme values

a) Look at the plots (esp. scatterplots) generated in the previous question. Which 3-4 countries
appear most extreme? Why do you consider them extreme?

- Cook Islands(COK), Papua New Guinea(PNG) and Samoa(SAM) appear as the most extreme in almost all races, they are extreme because they acheived the worst records by a far margin from other countries.

&nbsp;

One approach to measuring "extremism" is to look at the distance (needs to be deffined!) between
an observation and the sample mean vector, i.e. we look how far one is from the average. Such
a distance can be called an multivariate residual for the given observation.






b) The most common residual is the Euclidean distance between the observation and sample
mean vector, i.e.

$$d(\vec{x}, \hat{x}) = \sqrt{(\vec{x}-\hat{x})^T(\vec{x}-\hat{x})}$$
This distance can be immediately generalized to the $L^r$, r > 0 distance as

$$d_L(\vec{x}, \hat{x}) = (\sum_{i=1}^p \left\lvert \vec{x}-\hat{x} \right\rvert^r)^{1/r}$$
where p is the dimension of the observation (here p = 7). 

Compute the squared Euclidean distance (i.e. r = 2) of the observation from the sample
mean for all 55 countries using R's matrix operations. First center the raw data by the means
to get $\vec{x} - \bar{x}$ for each country. Then do a calculation with matrices that will result in a matrix
that has on its diagonal the requested squared distance for each country. Copy this diagonal
to a vector and report on the five most extreme countries. In this questions you MAY NOT
use any loops.


```{r echo=FALSE}
centered           <- apply(X[,-1], 2, function(x){abs(x-mean(x))})
rownames(centered) <- X[[1]]
mat                <- sqrt(tcrossprod(centered))
dist1               <- sort(diag(mat),decreasing = TRUE)
top_5              <- round(dist1[1:5],2)
# for some reason sqrt function didnt work in rmarkdown
# so we had to copy the values from r console
```

&nbsp;

PNG: 67.63 

COK: 59.62 

SAM: 38.52 

BER: 20.62 

GBR: 18.59 

&nbsp;

c) The different variables have different scales so it is possible that the distances can be dominated by some few variables. To avoid this we can use the squared distance

$$d_V^2(\vec{x}, \hat{x}) = (\vec{x}-\hat{x})^TV^{-1}(\vec{x}-\hat{x})$$

where **V** is a diagonal matrix with variances of the appropriate variables on the diagonal.
The effect, is that for each variable the squared distance is divided by its variance and we
have a scaled independent distance. 


It is simple to compute this measure by standardizing the raw data with both means (cen-
tring) and standard deviations (scaling), and then compute the Euclidean distance for the
normalized data. Carry out these computations and conclude which countries are the most
extreme ones. How do your conclusions compare with the unnormalized ones? 

```{r echo=FALSE}
var                <- matrix(0,7,7)
diag(var)          <- diag(cov(X[,-1]))
sqrd_dist          <- ((centered) %*% solve(var)) %*% t(centered)
dist2               <- (sort(diag(sqrd_dist),decreasing = TRUE))
top_5sq             <- round(dist2[1:5],2)
top_5sq
```

- This method does not treat the varaibales equally, thats why countries like USA and Singapore shown as extremes.

&nbsp;

d) The most common statistical distance is the Mahalanobis distance

$$d_M^2(\vec{x}, \hat{x}) = (\vec{x}-\hat{x})^TC^{-1}(\vec{x}-\hat{x})$$

where C is the sample covariance matrix calculated from the data. With this measure we also
use the relationships (covariances) between the variables (and not only the marginal variances
as $d_v$ (.,.) does). Compute the Mahalanobis distance, which countries are most extreme now?


```{r echo=FALSE}
C                  <- cov(X[,-1])
mah                <- ((centered) %*% solve(C)) %*% t(centered)
dist3              <- (sort(diag(mah),decreasing = TRUE))
top_5mh            <- round(dist3[1:5],2)
top_5mh
```




e) Compare the results in b)-d). Some of the countries are in the upper end with all the
measures and perhaps they can be classiffied as extreme. Discuss this. But also notice the
different measures give rather different results (how does Sweden behave?). Summarize this
graphically. Produce Czekanowski's diagram using e.g. the **RMaCzek** package. In case of
problems please describe them.


```{r echo=FALSE}
Sweden1 <- which(row.names(as.data.frame(dist1)) =='SWE')
Sweden2 <- which(row.names(as.data.frame(dist2)) =='SWE')
Sweden3 <- which(row.names(as.data.frame(dist3)) =='SWE')

swe            <- data.frame(c(Sweden1, Sweden2, Sweden3), 3,1)
row.names(swe) <- c("Euclidean", "Squared", "Mahalanobis")
colnames(swe)  <- c("SWE Rank")
swe
```


```{r include=FALSE}
library("RMaCzek")
```

```{r echo=FALSE}
rownames(X) <- X[[1]]
czek        <- czek_matrix(X[,-1])
plot(czek)
```



## References:

- https://www.gastonsanchez.com/visually-enforced/how-to/2014/01/15/Center-data-in-R/ 
- http://geog.uoregon.edu/bartlein/courses/geog495/lec18.html

&nbsp, 

## Appendix

```{r echo=TRUE, eval=FALSE}
library(ggplot2)
library(psych)


X <- read.table("D:/Workshop/R/Multivariate Statistics/Data/T1-9.DAT")
colnames(X) <- c("Country", "100m", "200m", "400m", "800m", "1500m", "3000m", "Marathon")

Minimum               <- matrix(0,7,1)
colnames(Minimum)     <- c("Min.")

Median                <- matrix(0,7,1)
colnames(Median)      <- c("Median")

Mean                  <- matrix(0,7,1)
colnames(Mean)        <- c("Mean")

S_Deviation           <- matrix(0,7,1)
colnames(S_Deviation) <- c("SD")

Maximum               <- matrix(0,7,1)
colnames(Maximum)     <- c("Max.")


M <- function(X,var) {
  for(i in 1:var) {
    Minimum[i,]     <- min(X[,i+1])
    Median[i,]      <- median(X[,i+1])
    Mean[i,]        <- mean(X[,i+1])
    S_Deviation[i,] <- sd(X[,i+1])
    Maximum[i,]     <- max(X[,i+1])
  }
  result            <- round(cbind(Minimum, Median, Mean, S_Deviation, Maximum),2)
  rownames(result)  <- c("100m", "200m", "400m", "800m", "1500m", "3000m", "Marathon")
  return(result)
}
knitr::kable(M(X = X, var = 7))


pairs.panels(X)


ggplot(X, aes(x=X$`100m`)) +
  geom_histogram(aes(y=..density..), binwidth=.15,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 100m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`200m`)) +
  geom_histogram(aes(y=..density..), binwidth=.3,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 200m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`400m`)) +
  geom_histogram(aes(y=..density..), binwidth=.9,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Seconds") +
  ggtitle("Record Distribution of the Women's 400m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`800m`)) +
  geom_histogram(aes(y=..density..), binwidth=.03,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 800m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`1500m`)) +
  geom_histogram(aes(y=..density..), binwidth=.12,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 1500m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=X$`3000m`)) +
  geom_histogram(aes(y=..density..), binwidth=.4,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's 3000m Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))

ggplot(X, aes(x=Marathon)) +
  geom_histogram(aes(y=..density..), binwidth=6,
                 colour="black", fill="#00539CFF", alpha=0.3, size = 0.1) + 
  geom_density(fill="#FFD662FF", color="#e9ecef", alpha=0.3) +
  xlab("Record in Minutes") +
  ggtitle("Record Distribution of the Women's Marathon Race") +
  theme(axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        title = element_text(size=8))



Corr = round(cor(X[,-1]),2)
cat("Correlation Matrix:\n")

knitr::kable(Corr)

cat("Covariance Matrix:\n")
Cov = round(cov(X[,-1]),2)
knitr::kable(Cov)

library(ggrepel)
library("gridExtra")

p1 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,3]))+
  xlab("100m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p2 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,4]))+
  xlab("100m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p3 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,5]))+
  xlab("100m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p4 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,6]))+
  xlab("100m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p5 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,7]))+
  xlab("100m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p6 <- ggplot(data = as.data.frame(X), aes(y = X[,2], x = X[,8]))+
  xlab("100m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

grid.arrange(p1, p2, p3, p4, p5, p6)



p7 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,2]))+
  xlab("200m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p8 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,4]))+
  xlab("200m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p9 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,5]))+
  xlab("200m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p10 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,6]))+
  xlab("200m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p11 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,7]))+
  xlab("200m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p12 <- ggplot(data = as.data.frame(X), aes(y = X[,3], x = X[,8]))+
  xlab("200m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p13 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,2]))+
  xlab("400m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p14 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,3]))+
  xlab("400m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p15 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,5]))+
  xlab("400m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p16 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,6]))+
  xlab("400m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p17 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,7]))+
  xlab("400m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p18 <- ggplot(data = as.data.frame(X), aes(y = X[,4], x = X[,8]))+
  xlab("400m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p19 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,2]))+
  xlab("800m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p20 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,3]))+
  xlab("800m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p21 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,4]))+
  xlab("800m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p22 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,6]))+
  xlab("800m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p23 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,7]))+
  xlab("800m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p24 <- ggplot(data = as.data.frame(X), aes(y = X[,5], x = X[,8]))+
  xlab("800m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p25 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,2]))+
  xlab("1500m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p26 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,3]))+
  xlab("1500m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p27 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,4]))+
  xlab("1500m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p28 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,5]))+
  xlab("1500m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p29 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,7]))+
  xlab("1500m") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p30 <- ggplot(data = as.data.frame(X), aes(y = X[,6], x = X[,8]))+
  xlab("1500m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p31 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,2]))+
  xlab("30000m") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p32 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,3]))+
  xlab("3000m") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p33 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,4]))+
  xlab("3000m") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p34 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,5]))+
  xlab("3000m") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p35 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,6]))+
  xlab("3000m") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p36 <- ggplot(data = as.data.frame(X), aes(y = X[,7], x = X[,8]))+
  xlab("3000m") + ylab("Marathon")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')

p37 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,2]))+
  xlab("Marathon") + ylab("100m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p38 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,3]))+
  xlab("Marathon") + ylab("200m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p39 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,4]))+
  xlab("Marathon") + ylab("400m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p40 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,5]))+
  xlab("Marathon") + ylab("800m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p41 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,6]))+
  xlab("Marathon") + ylab("1500m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red') 

p42 <- ggplot(data = as.data.frame(X), aes(y = X[,8], x = X[,7]))+
  xlab("Marathon") + ylab("3000m")+
  geom_text_repel(label = X[,1], size = 2)+
  geom_point(color = 'red')


grid.arrange(p7, p8, p9, p10, p11, p12)
grid.arrange(p13, p14, p15, p16, p17, p18) 
grid.arrange(p19, p20, p21, p22, p23, p24) 
grid.arrange(p25, p26, p27, p28, p29, p30) 
grid.arrange(p31, p32, p33, p34, p35, p36) 
grid.arrange(p37, p38, p39, p40, p41, p42) 

library(car)

par(mar= c(1,1,1,1))

scatterplotMatrix(X[,2:8])

library(corrplot)




corrplot(Corr, type = "upper", 
        tl.col = "black", tl.srt = 45)


centered           <- apply(X[,-1], 2, function(x){abs(x-mean(x))})
rownames(centered) <- X[[1]]
mat                <- sqrt(tcrossprod(centered))
dist1               <- sort(diag(mat),decreasing = TRUE)
top_5              <- round(dist1[1:5],2)
# for some reason sqrt function didnt work in rmarkdown
# so we had to copy the values from r console

var                <- matrix(0,7,7)
diag(var)          <- diag(cov(X[,-1]))
sqrd_dist          <- ((centered) %*% solve(var)) %*% t(centered)
dist2               <- (sort(diag(sqrd_dist),decreasing = TRUE))
top_5sq             <- round(dist2[1:5],2)
top_5sq

C                  <- cov(X[,-1])
mah                <- ((centered) %*% solve(C)) %*% t(centered)
dist3              <- (sort(diag(mah),decreasing = TRUE))
top_5mh            <- round(dist3[1:5],2)
top_5mh

Sweden1 <- which(row.names(as.data.frame(dist1)) =='SWE')
Sweden2 <- which(row.names(as.data.frame(dist2)) =='SWE')
Sweden3 <- which(row.names(as.data.frame(dist3)) =='SWE')

swe            <- data.frame(c(Sweden1, Sweden2, Sweden3), 3,1)
row.names(swe) <- c("Euclidean", "Squared", "Mahalanobis")
colnames(swe)  <- c("SWE Rank")
swe

library("RMaCzek")

rownames(X) <- X[[1]]
czek        <- czek_matrix(X[,-1])
plot(czek)

```




















