---
title: "Assignment 3: Principle Component and factor analysis"
author: "Mohsen Pirmoradiyan, Ahmed Alhasan, Asad Enver, Ali Etminan, Mubarak Hussain"
date: "2019/12/05"
output: pdf_document
---

```{r setup, include=FALSE}
library(corrplot)
library(ggplot2)
library(psych)
library(ggrepel)
library(gridExtra )
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1: Principal components, including interpretation of them


### (a) Obtain the sample correlation matrix R for these data, and determine its eigenvalues and eigenvectors.

Initially, We read the T1-9 data and add column names

```{r data}

newdata <- read.delim("D:/Machine Learning/Workshop/Multivariate Statistics/Data/T1-9.dat")
colnames(newdata) <- c("Country", "100m", "200m", "400m", "800m", "1500m", 
                       "3000m", "Marathon")

```

Here, we create the correlation matrix on the data with two decimal points and plot the correlation matrix for a visual comprehension. We use the *corrplot()* function from the *corrplot* package.

```{r cor, echo=FALSE, fig.height = 4.2 }

cor.mat <- round(cor(newdata[,-1]),2)
cor.mat 
corrplot(cor.mat)

```


Now using R's *eigen()* function we can generate the eigenvalues and vectors

```{r eigens}

eigens <- eigen(cor.mat)
eigens

```


### (b) Determine the first two principal components for the standardized variables. Prepare a table showing the correlations of the standardized variables with the components, and the cumulative percentage of the total (standardized) sample variance explained by the two components.


From part *(a)* we have the eigenvectors of the correlation matrix. Each eigenvector represents the principle component for a variable. Now we can generate a table showing the correlation between variables and different components.

```{r cortable}

eigen.vectors <- eigens$vectors
row.names(eigen.vectors) <- c("100m", "200m", "400m", "800m", "1500m",
                              "3000m", "Marathon")
colnames(eigen.vectors) <- c("COMP1", "COMP2", "COMP3", "COMP4", "COMP5", 
                             "COMP6", "COMP7")
eigen.vectors

```


Alternatively, we can use the *prcomp()* function to determine the components as well as their standard deviations. 

```{r prcomp}

alt.pca <- prcomp(newdata[,-1], center = TRUE, scale. = TRUE)
alt.pca

```

However, we are going to stick to the values obtained from part (a) for the rest of the solution.

The first two principle components:

```{r PCs}

eigens$vectors[,1:2]

```

As the next step, we draw a plot using the first principle component (PC1) for the x axis and the second principle component (PC2) for the y axis:

```{r PCplot, echo=FALSE}

ggplot() +
  geom_point(aes(alt.pca$x[,1],alt.pca$x[,2])) +
  xlab("PC1") + ylab("PC2") + ggtitle("Principle Components")

```

For the next part, we calculate the contribution (accounted sample variance) of the first principle component in percentage

```{r PC1}

PC1 <- (eigens$values[1]/sum(eigens$values))*100
PC1

```

We do the same calculations for the second principle component

```{r PC2}

PC2 <- (eigens$values[2]/sum(eigens$values))*100
PC2

```

Finally, we calculate the cumulative proportion of the total sample variance for PC1 and PC2 in percentage

```{r totalvar}

total.var <- PC1 + PC2
total.var

```

From the results we can see that the first two principle components have a 92.1% contribution to the total variance. Therefore, the cumulative proportion of the total sample variance for PC1 and PC2 is 0.92

### (c) Interpret the two principal components obtained in Part b. (Note that the first component is essentially a normalized unit vector and might measure the athletic excellence of a given nation. The second component might measure the relative strength of a nation at the various running distances.)

From the results we can see that all variables contribute almost equally to the first component. However, for the second component we can see a larger range in the values. This could be an indication of a nation's strength (running time) for various distances. We could consider the second component as the **"Distance Component"**

### (d) Rank the nations based on their score on the first principal component. Does this ranking correspond with your inituitive notion of athletic excellence for the various countries?

```{r}
records <- newdata[,-1]
rownames(records) <- newdata[,1]
adj <- records * alt.pca$rotation[,1]
rank <-apply(adj, 1, mean)
rank[order(rank)]
```

- The countries ranking according to the first PC corresponds with their overall performance across all races.










```{r}
test  <- function(records, m= 3){
  p   <- dim(records)[2]
  n   <- dim(records)[1]
  

  num <- tcrossprod(loadings) + diag(pc$uniquenesses)
  T   <- (n-1-(2*p+4*m+5)/6) * log(det(num)) / det((n-1)/n*S)
  T0  <- qchisq(0.95, df = (((p-m)**2-p-m)/2))
  
  if(T>T0){
    print("Reject H0")
    }
  else{
    print("Can not reject H0")
    }
}


S  <- cov(records)
pc <- principal(S, nfactors=3, rotate="varimax")
loadings <- pc$loadings[,1:3]
test(records, 3)

s_scores  <-as.matrix(records)%*%loadings
```

```{r fig.align='center', fig.height=7}
p1 <- ggplot(data = as.data.frame(s_scores), aes(y = s_scores[,1], x = seq(1, length(s_scores[,1]))))+
        geom_text_repel(label = rownames(s_scores), size = 2)+
        xlab("Country Index")+
        ylab("Scores")+
        ggtitle("RC2")+
        geom_point(color = 'red') 

p2 <- ggplot(data = as.data.frame(s_scores), aes(y = s_scores[,2], x = seq(1, length(s_scores[,2]))))+
        geom_text_repel(label = rownames(s_scores), size = 2)+
        xlab("Country Index")+
        ylab("Scores")+
        ggtitle("RC1")+
        geom_point(color = 'red') 

p3 <- ggplot(data = as.data.frame(s_scores), aes(y = s_scores[,3], x = seq(1, length(s_scores[,3]))))+
        geom_text_repel(label = rownames(s_scores), size = 2)+
        xlab("Country Index")+
        ylab("Scores")+
        ggtitle("RC3")+
        geom_point(color = 'red') 


grid.arrange(p1, p2, p3, nrow = 3)

```


# Appendix

```{r appendix, eval=FALSE}


library(corrplot)
library(ggplot2)

newdata <- read.delim("D:/Machine Learning/Workshop/Multivariate Statistics/Data/T1-9.dat")
colnames(newdata) <- c("Country", "100m", "200m", "400m", "800m", "1500m", 
                       "3000m", "Marathon")

cor.mat <- round(cor(newdata[,-1]),2)
cor.mat 

corrplot(cor.mat)

eigens <- eigen(cor.mat)
eigens


eigen.vectors <- eigens$vectors
row.names(eigen.vectors) <- c("100m", "200m", "400m", "800m", "1500m", 
                              "3000m", "Marathon")
colnames(eigen.vectors) <- c("COMP1", "COMP2", "COMP3", "COMP4", "COMP5",
                             "COMP6", "COMP7")
eigen.vectors


alt.pca <- prcomp(newdata[,-1], center = TRUE, scale. = TRUE)
alt.pca

eigens$vectors[,1:2]
prcomp(cor.mat)


ggplot() +
  geom_point(aes(alt.pca$x[,1],alt.pca$x[,2])) +
  xlab("PC1") + ylab("PC2") + ggtitle("Principle Components")

PC1 <- (eigens$values[1]/sum(eigens$values))*100
PC1

PC2 <- (eigens$values[2]/sum(eigens$values))*100
PC2

total.var <- PC1 + PC2
total.var

```